Active Learning from Blackbox to Timed Connectors

### 2016-01-20 paper/paper.tex
.

In this paper we address this question by proposing an active learning framework as shown in Figure
\ref{fig:howto} to automatically extract timed connector models from blackbox implementations. 
To achieve this goal, we first introduce \emph{parameterized Mealy machine} as a parameterized
semantics for timed Reo connectors, which can be transformed to concrete Mealy machines with a
given alphabet. Then the $L*$ algorithm \cite{DBLP:journals/iandc/Angluin87} is adapted and
optimized to extract Mealy machines with timed action from the connector implementations as
blackboxes.

@@

Figure \ref{fig:basic} shows the graphical representation of some basic channel types in Reo, whose
behavior are described as follows:

@@

Channels are joint together in \emph{nodes}. There are three types of nodes:
\emph{source}, \emph{sink} and \emph{mixed} nodes, depending on whether all channel ends that
coincide on a node are source ends, sink ends or a combination of both.
(see Figure \ref{fig:node})

@@

Figure \ref{fig:reoconnector} provides a simple example to illustrate how complex
connectors are constructed and used in coordination. In this example, we have three components
COM1,COM2 and MONITOR. With an \emph{alternator} connector as shown in Figure
\ref{fig:reoconnector}, MONITOR can obtain data items from COM1 and COM2 alternately. This example
has been implemented in \texttt{Golang} as shown in Section \ref{sec:reolib}.
Besides, all the connectors are supposed to have \emph{deterministic} behavior hereinafter to meet
the requirement of active learning algorithm L*.

@@

Figure \ref{fig:activelearning} shows the sketch of active learning, wherein:
\begin{itemize}
  \item[-] \emph{Teacher} and \emph{Student}: Active learning is an interactive process where
    students ask questions and teachers answer. Here learning algorithm plays the role of student.
  \item[-] \emph{Oracles} is an interface specifying which kind of questions can be answered by the
    teacher.
  \item[-] \emph{SUL} is an abbreviation of System Under Learn. In this paper, we take blackbox
    models as our SULs.
  \item[-] \emph{CEGAR} indicates Counter-Example Guided Abstraction
    Refinement\cite{DBLP:conf/cav/ClarkeGJLV00}. In active learning, we need counter-examples to
    guide us on further queries and cover the undistinguished states.
\end{itemize}

@@

These queries are given by \emph{learning algorithms}, or so-called \emph{students} in Figure
\ref{fig:activelearning}. From the \emph{mq} results, a learning algorithm constructs a
\emph{hypothesis} and then check it with \emph{eq}. If counter-examples are found, we turn back and
repeat the hypothesis construction until the equivalence query returns \emph{true}.

@@

Besides, the concrete Mealy machine (where $\Sigma=\{d\}$) is shown in Figure \ref{fig:pmmfifo},
where labels of edges are in form of $\frac{input}{output}$. All $\rblock$ edges are ignored for
clearance.
\begin{figure}[ht]
  \begin{center}
    \input{./images/pmmfifo.tex}
  \end{center}
  \caption{PMM-based Semantics of $\smap{FIFO1}_\Sigma$, where $\Sigma=\{d\}$}
  \label{fig:pmmfifo}
\end{figure}

@@

As mentioned above, we cannot connect a sink end of $\mathcal{PM}_2$ to a source end of
$\mathcal{PM}_1$, which makes it difficult to construct connectors like alternator in Figure
\ref{fig:reoconnector}. To solve the problem, we define a \emph{link} operator to connect sink
ends and source ends in the same connector.

@@

Formally, an observation table (see Figure \ref{fig:hypo} as an example) is determined by a tuple $(S,D)$ where
$S\subset\mathcal{I}^*$ is a set of the access sequences of states and $D\subset\mathcal{I}^*$
is a set of suffixes. In such tables, columns are labelled with suffixes in $D$, and rows are
labelled with input sequences. As shown in Figure \ref{fig:hypo}, rows are divided into two parts. The upper
part, labelled with access sequence $s\in S$, denotes the states covered in the current hypothesis.
The lower part, labelled with sequences in $\{sa|s\in S,a\in\mathcal{I}\}\backslash S$ (representing
any potentially unclosed transition) denotes all the transitions outside $s\in S$. A cell with row
label $s$ and column label $d$ is filled with $mq(sd)$.

@@

Firstly, $obs.S$ is initialized with the initial state (denoted by $\varepsilon$), and $obs.D$ is
initialized with all one-step suffixes.
Then we explore all the access sequences in $obs.S$ and calculate its successors. Here
$unclosed$ consists of access sequences that has no equivalent sequence in $obs.S$. 
For example, Figure \ref{fig:hypo} shows the $obs$ of a 2-Timer channel after the first
iteration. The five successors of $\varepsilon$ are presented at the bottom of the table, where
four are equivalent with $\varepsilon$ but $(a,\rnoread)$ is not. Therefore, we take $(a,\rnoread)$ as a
brand-new state and all of its successors need further exploration.

@@

\begin{figure}[ht]
  \begin{center}
    \begin{displaymath}
      \begin{array}{l||ccccc||c}
        \hline
        & (a,\rread) & (\bot,\rread) & (a,\rnoread) & (\bot,\rnoread) & T & acc\\
        \hline\hline
        \varepsilon & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon \\
        (a,\rnoread) & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        \hline
        (a,\rread) & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon\\
        (\bot,\rread) & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon \\
        (\bot,\rnoread) & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon \\
        T & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon \\
        (a,\rnoread),(a,\rread) & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        (a,\rnoread),(\bot,\rread) & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        (a,\rnoread),(a,\rnoread) & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        (a,\rnoread),(\bot,\rnoread) & \rblock & \rblock & \rblock & \rempty & \rempty &
        (a,\rnoread) \\
        (a,\rnoread),T & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        \hline
      \end{array}
    \end{displaymath}
  \end{center}
  \caption{A Closed Version of Figure \ref{fig:hypo}}
  \label{fig:hypo2}
\end{figure}

\subsection{Counter Examples' Analysis} 
Apparently, the closed hypothesis presented in Figure \ref{fig:hypo2} is different with the 2-Timer
channel. It's easy to find a counter-example $s_0=(a,\rnoread),T,T,(\bot,\rread)$ where $mq(s_0)=a$ while
according to the hypothesis, the result should be $\rempty$. In this section, we show how to
find and analyze counter-examples using the method in \cite{DBLP:conf/sfm/SteffenHM11}.

@@

Figure \ref{fig:expfifosemantics} shows the learning result of this example where $n=2$ and
$\Sigma=\{a\}$.
To simplify the graph, we ignore all the trivial transitions $\frac{(\bot,\rnoread)}{\rempty}$
and block transitions. More details of this case can be found in our \emph{github repo}.

@@

Our implementation in \texttt{Golang} is well-prepared not only for academic use
but also for practical concurrent programming. The following code shows how to construct an alternator
connector (see Figure \ref{fig:reoconnector}) in \texttt{Golang}. An intact version of this example
can be found in our github repo.

>>>>

In this paper we address this question by proposing an active learning framework as shown in Fig.
\ref{fig:howto} to automatically extract timed connector models from blackbox implementations. 
To achieve this goal, we first introduce \emph{parameterized Mealy machine} as a parameterized
semantics for timed Reo connectors, which can be transformed to concrete Mealy machines with a
given alphabet. Then the $L*$ algorithm \cite{DBLP:journals/iandc/Angluin87} is adapted and
optimized to extract Mealy machines with timed action from the connector implementations as
blackboxes.

@@

Fig. \ref{fig:basic} shows the graphical representation of some basic channel types in Reo, whose
behavior are described as follows:

@@

Channels are joint together in \emph{nodes}. There are three types of nodes:
\emph{source}, \emph{sink} and \emph{mixed} nodes, depending on whether all channel ends that
coincide on a node are source ends, sink ends or a combination of both.
(see Fig. \ref{fig:node})

@@

Fig. \ref{fig:reoconnector} provides a simple example to illustrate how complex
connectors are constructed and used in coordination. In this example, we have three components
COM1,COM2 and MONITOR. With an \emph{alternator} connector as shown in Fig.
\ref{fig:reoconnector}, MONITOR can obtain data items from COM1 and COM2 alternately. This example
has been implemented in \texttt{Golang} as shown in Section \ref{sec:reolib}.
Besides, all the connectors are supposed to have \emph{deterministic} behavior hereinafter to meet
the requirement of active learning algorithm L*.

@@

Fig. \ref{fig:activelearning} shows the sketch of active learning, wherein:
\begin{itemize}
  \item[-] \emph{Teacher} and \emph{Student}: Active learning is an interactive process where
    students ask questions and teachers answer. Here learning algorithm plays the role of student.
  \item[-] \emph{Oracles} is an interface specifying which kind of questions can be answered by the
    teacher.
  \item[-] \emph{SUL} is an abbreviation of System Under Learn. In this paper, we take blackbox
    models as our SULs.
  \item[-] \emph{CEGAR} indicates Counter-Example Guided Abstraction
    Refinement\cite{DBLP:conf/cav/ClarkeGJLV00}. In active learning, we need counter-examples to
    guide us on further queries and cover the undistinguished states.
\end{itemize}

@@

These queries are given by \emph{learning algorithms}, or so-called \emph{students} in Fig.
\ref{fig:activelearning}. From the \emph{mq} results, a learning algorithm constructs a
\emph{hypothesis} and then check it with \emph{eq}. If counter-examples are found, we turn back and
repeat the hypothesis construction until the equivalence query returns \emph{true}.

@@

Besides, the concrete Mealy machine (where $\Sigma=\{d\}$) is shown in Fig. \ref{fig:pmmfifo},
where labels of edges are in form of $\frac{input}{output}$. All $\rblock$ edges are ignored for
clearance.
\begin{figure}[ht]
  \begin{center}
    \input{./images/pmmfifo.tex}
  \end{center}
  \caption{PMM-based Semantics of $\smap{FIFO1}_\Sigma$, where $\Sigma=\{d\}$}
  \label{fig:pmmfifo}
\end{figure}

@@

As mentioned above, we cannot connect a sink end of $\mathcal{PM}_2$ to a source end of
$\mathcal{PM}_1$, which makes it difficult to construct connectors like alternator in Fig.
\ref{fig:reoconnector}. To solve the problem, we define a \emph{link} operator to connect sink
ends and source ends in the same connector.

@@

Formally, an observation table (see Fig. \ref{fig:hypo} as an example) is determined by a tuple $(S,D)$ where
$S\subset\mathcal{I}^*$ is a set of the access sequences of states and $D\subset\mathcal{I}^*$
is a set of suffixes. In such tables, columns are labelled with suffixes in $D$, and rows are
labelled with input sequences. As shown in Fig. \ref{fig:hypo}, rows are divided into two parts. The upper
part, labelled with access sequence $s\in S$, denotes the states covered in the current hypothesis.
The lower part, labelled with sequences in $\{sa|s\in S,a\in\mathcal{I}\}\backslash S$ (representing
any potentially unclosed transition) denotes all the transitions outside $s\in S$. A cell with row
label $s$ and column label $d$ is filled with $mq(sd)$.

@@

Firstly, $obs.S$ is initialized with the initial state (denoted by $\varepsilon$), and $obs.D$ is
initialized with all one-step suffixes.
Then we explore all the access sequences in $obs.S$ and calculate its successors. Here
$unclosed$ consists of access sequences that has no equivalent sequence in $obs.S$. 
For example, Fig. \ref{fig:hypo} shows the $obs$ of a 2-Timer channel after the first
iteration. The five successors of $\varepsilon$ are presented at the bottom of the table, where
four are equivalent with $\varepsilon$ but $(a,\rnoread)$ is not. Therefore, we take $(a,\rnoread)$ as a
brand-new state and all of its successors need further exploration.

@@

\begin{figure}[ht]
  \begin{center}
    \begin{displaymath}
      \begin{array}{l||ccccc||c}
        \hline
        & (a,\rread) & (\bot,\rread) & (a,\rnoread) & (\bot,\rnoread) & T & acc\\
        \hline\hline
        \varepsilon & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon \\
        (a,\rnoread) & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        \hline
        (a,\rread) & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon\\
        (\bot,\rread) & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon \\
        (\bot,\rnoread) & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon \\
        T & \rblock & \rblock & \rempty & \rempty & \rempty & \varepsilon \\
        (a,\rnoread),(a,\rread) & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        (a,\rnoread),(\bot,\rread) & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        (a,\rnoread),(a,\rnoread) & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        (a,\rnoread),(\bot,\rnoread) & \rblock & \rblock & \rblock & \rempty & \rempty &
        (a,\rnoread) \\
        (a,\rnoread),T & \rblock & \rblock & \rblock & \rempty & \rempty & (a,\rnoread) \\
        \hline
      \end{array}
    \end{displaymath}
  \end{center}
  \caption{A Closed Version of Fig. \ref{fig:hypo}}
  \label{fig:hypo2}
\end{figure}

\subsection{Counter Examples' Analysis} 
Apparently, the closed hypothesis presented in Fig. \ref{fig:hypo2} is different with the 2-Timer
channel. It's easy to find a counter-example $s_0=(a,\rnoread),T,T,(\bot,\rread)$ where $mq(s_0)=a$ while
according to the hypothesis, the result should be $\rempty$. In this section, we show how to
find and analyze counter-examples using the method in \cite{DBLP:conf/sfm/SteffenHM11}.

@@

Fig. \ref{fig:expfifosemantics} shows the learning result of this example where $n=2$ and
$\Sigma=\{a\}$.
To simplify the graph, we ignore all the trivial transitions $\frac{(\bot,\rnoread)}{\rempty}$
and block transitions. More details of this case can be found in our \emph{github repo}.

@@

Our implementation in \texttt{Golang} is well-prepared not only for academic use
but also for practical concurrent programming. The following code shows how to construct an alternator
connector (see Fig. \ref{fig:reoconnector}) in \texttt{Golang}. An intact version of this example
can be found in our github repo.
